### 2019-03-06 writeup/paper.tex
can load files now

\documentclass{article}

\usepackage[english]{babel}
\usepackage[square, numbers]{natbib}
\usepackage{url}
\bibliographystyle{abbrvnat}
\setcitestyle{authoryear,open={(},close={)}}

\usepackage[autostyle, english = american]{csquotes}
\MakeOuterQuote{"}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{gensymb}
\usepackage{verbatim}
\usepackage{amsfonts}
\usepackage{mathrsfs}
\usepackage{tikz-cd}
\usepackage{graphicx}
\usepackage{microtype}
\usepackage{centernot}
\usepackage{mathtools}
\usepackage[margin=1.0in]{geometry}
\usepackage{algorithm, algpseudocode}
\usepackage{bm}
\usepackage{caption}

\usepackage{pgfplots}
\pgfplotsset{compat=newest}
\usepackage{tikz}
\usetikzlibrary{shapes}

\usepackage[title]{appendix}
\usepackage{cleveref}

\pgfplotsset{height=8cm, width=15cm,compat=1.9}

\DeclareMathOperator*{\esssup}{ess\,sup}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}
\newcommand{\Ex}{\mathbb{E}}
\newcommand*\diff{\mathop{}\!\mathrm{d}}
\newcommand*\Diff[1]{\mathop{}\!\mathrm{d^#1}}

\newtheorem{definition}{Definition}
\newtheorem{exercise}{Exercise}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}{Proposition}
\newtheorem{example}{Example}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}

\title{CMPUT 609 Course Project}
\author{Alan Chan}
\date{}

\begin{document}
  \maketitle

  \section{Introduction}
  Policy gradient methods in Reinforcement Learning (RL) are a way of learning a policy in a Markov Decision Process (MDP) so as to maximize an objective function, usually the expected reward.

  \section{Methods}

  \section{Results & Discussion}

  \section{Conclusion}
\end{document}

### 2019-03-06 main.py

@@ -9,15 +9,15 @@ envs = [env1, env2, env3]
 np.random.seed(609)
 
 load_data = False
-run_experiment = False
-test_run = True
-save_data = False
-episodes = 50
+run_experiment = True
+test_run = False
+save_data = True
+episodes = 100
 gamma = 0.99
 
 # learning rates we will try
-lrs = [1e-4 * (2 ** i) for i in range(10)]
-trials = 2
+lrs = [1e-4 * (2 ** i) for i in range(5)]
+trials = 10
 
 returns = {"env": [], "lr": [], "use_entropy": [], "episode": [], "return": []}
 episode_ixs = [i + 1 for i in range(episodes)]
@@ -58,17 +58,25 @@ if run_experiment == True:
                     #returns["ts"] = np.average(data, axis = 0)
                     #returns["std"] = np.std(data, axis = 0)
 
+if load_data == True:
+    df = pd.read_csv("experiments.csv")
+else:
+    df = pd.DataFrame(returns)
 ## plot data
-# plots needed: Best lr curves for each alg, plot of all lr's within algs
-df = pd.DataFrame(returns)
-# best lr's within alg
-for use_entropy in (True, False):
-    ax = seaborn.lineplot(x = "episode", y = "return", hue = "lr", legend = "full", data = df.loc[df.use_entropy == use_entropy, :])
-    plt.savefig("{}-experiments.png")
-    plt.show()
+# plots needed: Best lr curves for each alg and env, plot of all lr's within algs for each env
+# best lr's within alg for each env
+for env in envs:
+    for use_entropy in (True, False):
+        env_name = env.unwrapped.spec.id
+        curr_env = df.env == env_name
+        curr_entropy = df.use_entropy == use_entropy
+        ax = seaborn.lineplot(x = "episode", y = "return", hue = "lr", legend = "full", data = df.loc[curr_env & curr_entropy, :])
+        plt.savefig("figs/{}-{}-experiments.png".format(env, curr_entropy))
+        plt.show()
 
 # save the data
 if save_data is True:
     df.to_csv("experiments.csv", index = False)
 
 ## do statistical tests?
+# other ways of evaluating: episodes required to get to a certain reward

### 2019-03-07 includes.py
fixed dataframe issue that prevented conversion of data to dataframe

@@ -7,5 +7,6 @@ from torch.distributions import Categorical
 import seaborn
 from matplotlib import pyplot as plt
 import pandas as pd
+import time
 
 device = torch.device('cuda' if torch.cuda.is_available() else 'cpu')

### 2019-03-07 main.py

@@ -4,7 +4,7 @@ env1 = gym.make('CartPole-v0')
 env2 = gym.make('MountainCar-v0')
 env3 = gym.make("Acrobot-v1")
 
-envs = [env1, env2, env3]
+envs = [env1]
 
 np.random.seed(609)
 
@@ -45,18 +45,23 @@ if run_experiment == True:
     for env in envs:
         for lr in lrs:
             for use_entropy in (True, False):
+                t_start = time.time()
                 env_name = env.unwrapped.spec.id
                 print("performing trials for env = {}, lr = {}, use_entropy = {}".format(env_name, lr, use_entropy))
                 for trial in range(trials):
-                    returns["env"].append(env_name)
-                    returns["lr"].append(lr)
-                    returns["use_entropy"].append(use_entropy)
-                    # add index to mark the episodes in the returns dict?
+                    returns["env"] += [env_name] * episodes
+                    returns["lr"] += [lr] * episodes
+                    returns["use_entropy"] += [use_entropy] * episodes
+                    trial_t_start = time.time()
                     data = train(env = env, lr = lr, gamma = gamma, use_entropy = use_entropy, episodes = episodes)
+                    trial_t_end = time.time()
                     returns["episode"] += episode_ixs
                     returns["return"] += data
+                    print("trial {} took {}s".format(trial + 1, trial_t_end - trial_t_start))
                     #returns["ts"] = np.average(data, axis = 0)
                     #returns["std"] = np.std(data, axis = 0)
+                t_end = time.time()
+                print("time to complete: {}s".format(t_end - t_start))
 
 if load_data == True:
     df = pd.read_csv("experiments.csv")

### 2019-03-07 main.py
fixed plotting name issues

@@ -11,13 +11,13 @@ np.random.seed(609)
 load_data = False
 run_experiment = True
 test_run = False
-save_data = True
+save_data = not(load_data)
 episodes = 100
 gamma = 0.99
 
 # learning rates we will try
 lrs = [1e-4 * (2 ** i) for i in range(5)]
-trials = 10
+trials = 20
 
 returns = {"env": [], "lr": [], "use_entropy": [], "episode": [], "return": []}
 episode_ixs = [i + 1 for i in range(episodes)]
@@ -65,8 +65,15 @@ if run_experiment == True:
 
 if load_data == True:
     df = pd.read_csv("experiments.csv")
+    print("data loaded")
 else:
     df = pd.DataFrame(returns)
+
+# save the data
+if save_data is True:
+    df.to_csv("experiments.csv", index = False)
+    print("data saved")
+
 ## plot data
 # plots needed: Best lr curves for each alg and env, plot of all lr's within algs for each env
 # best lr's within alg for each env
@@ -76,12 +83,10 @@ for env in envs:
         curr_env = df.env == env_name
         curr_entropy = df.use_entropy == use_entropy
         ax = seaborn.lineplot(x = "episode", y = "return", hue = "lr", legend = "full", data = df.loc[curr_env & curr_entropy, :])
-        plt.savefig("figs/{}-{}-experiments.png".format(env, curr_entropy))
+        ax.set_title("Learning curves for use_entropy = {} on {}".format(str(use_entropy), env_name))
+        plt.savefig("figs\\{}-{}-experiments.png".format(env_name, str(use_entropy)))
         plt.show()
 
-# save the data
-if save_data is True:
-    df.to_csv("experiments.csv", index = False)
 
 ## do statistical tests?
 # other ways of evaluating: episodes required to get to a certain reward

### 2019-03-07 main.py
moved random seed to make comparisons fair

@@ -6,8 +6,6 @@ env3 = gym.make("Acrobot-v1")
 
 envs = [env1]
 
-np.random.seed(609)
-
 load_data = False
 run_experiment = True
 test_run = False
@@ -45,6 +43,7 @@ if run_experiment == True:
     for env in envs:
         for lr in lrs:
             for use_entropy in (True, False):
+                np.random.seed(609) # the random seed should be reset for every type of experiment since we want fair comparisons between algs
                 t_start = time.time()
                 env_name = env.unwrapped.spec.id
                 print("performing trials for env = {}, lr = {}, use_entropy = {}".format(env_name, lr, use_entropy))

### 2019-03-08 algs.py
added way of limiting episode length for some environments

@@ -19,7 +19,7 @@ class Policy(nn.Module):
             probs = self.softmax(self.layer(x)).squeeze().tolist()
             return -sum([p * math.log(p) for p in probs])
 
-def train(env, lr, gamma, use_entropy, episodes = 100):
+def train(env, lr, gamma, use_entropy, episodes = 100, episode_length = None):
     # do REINFORCE because we only have one step-size
     # can do actor-critic later if there is time
     action_dim = env.action_space.n
@@ -41,6 +41,7 @@ def train(env, lr, gamma, use_entropy, episodes = 100):
 
         states = [torch.Tensor(obs).to(device)]
         actions = []
+        t = 0 # to index episode time
         while done is not True:
             #env.render()
 
@@ -59,6 +60,11 @@ def train(env, lr, gamma, use_entropy, episodes = 100):
             states.append(torch.Tensor(obs).to(device))
             rewards.append(reward)
             actions.append(action.item())
+
+            t += 1
+
+            if episode_length != None and t >= episode_length:
+                break
         for i in range(len(actions)):
             if i > len(rewards) - 1:
                 G = 0

### 2019-03-08 main.py

@@ -3,6 +3,11 @@ from algs import *
 env1 = gym.make('CartPole-v0')
 env2 = gym.make('MountainCar-v0')
 env3 = gym.make("Acrobot-v1")
+env4 = gym.make("LunarLander-v2")
+
+# env1 and env4 have episodes that are short length so don't have to artificially terminate them
+
+# consider artificially terminating episodes if they run too long?
 
 envs = [env1]
 
@@ -12,6 +17,7 @@ test_run = False
 save_data = not(load_data)
 episodes = 100
 gamma = 0.99
+episode_length = 1000
 
 # learning rates we will try
 lrs = [1e-4 * (2 ** i) for i in range(5)]
@@ -51,9 +57,17 @@ if run_experiment == True:
                     returns["env"] += [env_name] * episodes
                     returns["lr"] += [lr] * episodes
                     returns["use_entropy"] += [use_entropy] * episodes
+
                     trial_t_start = time.time()
-                    data = train(env = env, lr = lr, gamma = gamma, use_entropy = use_entropy, episodes = episodes)
+
+                    if env == env3 or env == env4:
+                        # if we are doing either of these environments, limit the episode length b/c they take a v long time to run
+                        data = train(env = env, lr = lr, gamma = gamma, use_entropy = use_entropy, episodes = episodes, episode_length = episode_length)
+                    else:
+                        data = train(env = env, lr = lr, gamma = gamma, use_entropy = use_entropy, episodes = episodes)
+
                     trial_t_end = time.time()
+
                     returns["episode"] += episode_ixs
                     returns["return"] += data
                     print("trial {} took {}s".format(trial + 1, trial_t_end - trial_t_start))

### 2019-03-08 main.py
removed env that had dependency problem

@@ -3,13 +3,13 @@ from algs import *
 env1 = gym.make('CartPole-v0')
 env2 = gym.make('MountainCar-v0')
 env3 = gym.make("Acrobot-v1")
-env4 = gym.make("LunarLander-v2")
+#env4 = gym.make("LunarLander-v2")
 
 # env1 and env4 have episodes that are short length so don't have to artificially terminate them
 
 # consider artificially terminating episodes if they run too long?
 
-envs = [env1]
+envs = [env2]
 
 load_data = False
 run_experiment = True
@@ -60,7 +60,7 @@ if run_experiment == True:
 
                     trial_t_start = time.time()
 
-                    if env == env3 or env == env4:
+                    if env == env2 or env == env3:
                         # if we are doing either of these environments, limit the episode length b/c they take a v long time to run
                         data = train(env = env, lr = lr, gamma = gamma, use_entropy = use_entropy, episodes = episodes, episode_length = episode_length)
                     else:

### 2019-03-15 algs.py
fixed mountain car premature termination; added information content pseudoreward; general cleanup

@@ -13,7 +13,7 @@ class Policy(nn.Module):
         #assert torch.sum(torch.isnan(out) == 0), "policy output is nan, {}".format(self.layer.weight)
         return out
 
-def train(env, lr, gamma, use_entropy, episodes = 100, episode_length = None):
+def train(env, lr, gamma, pseudoreward, episodes = 100, episode_length = None, render_env = False, print_return = False):
     # do REINFORCE because we only have one step-size
     # can do actor-critic later if there is time
     action_dim = env.action_space.n
@@ -37,23 +37,31 @@ def train(env, lr, gamma, use_entropy, episodes = 100, episode_length = None):
         actions = []
         t = 0 # to index episode time
         while done is not True:
-            #env.render()
+            if render_env is True:
+                env.render()
 
             log_probs = policy(torch.Tensor(obs).to(device))
             m = Categorical(logits = log_probs)
             #print("time = {}".format(t), log_probs, obs)
             action = m.sample()
-            entropy = m.entropy()
 
             obs, reward, done, info = env.step(action.item())
+
+            # the real rewards we will use for evaluating the algorithms
             actual_rewards.append(reward)
-            # also try adding just -\ln p_i for the action that is taken
-            if use_entropy == True:
-                reward += entropy
 
             states.append(torch.Tensor(obs).to(device))
-            rewards.append(reward)
             actions.append(action.item())
+
+            # adding pseudorewards possibly to help with exploration
+            if pseudoreward == "entropy":
+                reward += m.entropy()
+            elif pseudoreward == "information_content":
+                reward -= m.log_prob(action)
+
+            # the pseudoreward where we add entropy or information content
+            rewards.append(reward)
+
 
             t += 1
 
@@ -78,5 +86,8 @@ def train(env, lr, gamma, use_entropy, episodes = 100, episode_length = None):
                 loss.backward(retain_graph = True)
             sgd.step()
         # calculate total return for this episode
-        returns.append(sum([(gamma ** i) * actual_rewards[i] for i in range(len(actual_rewards))]))
+        curr_return = sum([(gamma ** i) * actual_rewards[i] for i in range(len(actual_rewards))])
+        if print_return is True:
+            print("return on episode {} is {}".format(episode + 1, curr_return))
+        returns.append(curr_return)
     return returns

### 2019-03-15 main.py

@@ -1,7 +1,8 @@
 from algs import *
+from test_envs import *
 
 env1 = gym.make('CartPole-v0')
-env2 = gym.make('MountainCar-v0')
+env2 = gym.make('MountainCar-v0').env
 env3 = gym.make("Acrobot-v1")
 env4 = gym.make("LunarLander-v2")
 
@@ -9,31 +10,35 @@ env4 = gym.make("LunarLander-v2")
 
 # consider artificially terminating episodes if they run too long?
 
-envs = [env4, env1]
+envs = [env1]
+pseudorewards = ["entropy", "information_content"]
 
+run_experiment = False
+render_env = True
+print_return = True
 load_data = False
-run_experiment = True
-test_run = False
-save_data = not(load_data)
-episodes = 50
+test_run = not(run_experiment)
+save_data = run_experiment
+episodes = 40
 gamma = 0.99
 episode_length = None
 
 # learning rates we will try
 lrs = [1e-4 * (2 ** i) for i in range(5)]
-trials = 50
+trials = 2
 
-returns = {"env": [], "lr": [], "use_entropy": [], "episode": [], "return": []}
+returns = {"env": [], "lr": [], "pseudoreward": [], "episode": [], "return": []}
 episode_ixs = [i + 1 for i in range(episodes)]
 
 if test_run == True:
-    df = {"episode": [], "return": [], "use_entropy": [], "lr": []}
-    for lr in (1e-4, 2e-4, 4e-4):
-        for use_entropy in (True, False):
+    df = {"episode": [], "return": [], "pseudoreward": [], "lr": []}
+    for lr in (2e-4, 4e-4):
+        for pseudoreward in pseudorewards:
+            print("trials for pseudoreward = {}".format(pseudoreward))
             for trial in range(trials):
                 print("Starting trial {}".format(trial + 1))
-                data = train(env = env1, lr = lr, gamma = gamma, use_entropy = use_entropy, episodes = episodes)
-                df["use_entropy"] += [use_entropy] * episodes
+                data = train(env = env3, lr = lr, gamma = gamma, pseudoreward = pseudoreward, episodes = episodes, render_env = render_env, print_return = print_return, episode_length = episode_length)
+                df["pseudoreward"] += [pseudoreward] * episodes
                 df["return"] += data
                 df["episode"] += episode_ixs
                 df["lr"] += [lr] * episodes
@@ -48,7 +53,7 @@ if test_run == True:
 if run_experiment == True:
[…]

### 2019-03-15 main.py
removed artificial termination in acrobot

@@ -3,7 +3,7 @@ from test_envs import *
 
 env1 = gym.make('CartPole-v0')
 env2 = gym.make('MountainCar-v0').env
-env3 = gym.make("Acrobot-v1")
+env3 = gym.make("Acrobot-v1").env
 env4 = gym.make("LunarLander-v2")
 
 # env1 and env4 have episodes that are short length so don't have to artificially terminate them
@@ -37,7 +37,7 @@ if test_run == True:
             print("trials for pseudoreward = {}".format(pseudoreward))
             for trial in range(trials):
                 print("Starting trial {}".format(trial + 1))
-                data = train(env = env3, lr = lr, gamma = gamma, pseudoreward = pseudoreward, episodes = episodes, render_env = render_env, print_return = print_return, episode_length = episode_length)
+                data = train(env = env2, lr = lr, gamma = gamma, pseudoreward = pseudoreward, episodes = episodes, render_env = render_env, print_return = print_return, episode_length = episode_length)
                 df["pseudoreward"] += [pseudoreward] * episodes
                 df["return"] += data
                 df["episode"] += episode_ixs

### 2019-03-18 main.py
fixed plotting column issue in test runs

@@ -11,7 +11,7 @@ env4 = gym.make("LunarLander-v2")
 # consider artificially terminating episodes if they run too long?
 
 envs = [env1]
-pseudorewards = ["entropy", "information_content"]
+pseudorewards = ["none", "entropy", "information_content"]
 
 run_experiment = False
 render_env = True
@@ -19,13 +19,13 @@ print_return = True
 load_data = False
 test_run = not(run_experiment)
 save_data = run_experiment
-episodes = 40
+episodes = 3
 gamma = 0.99
-episode_length = None
+episode_length = 500
 
 # learning rates we will try
 lrs = [1e-4 * (2 ** i) for i in range(5)]
-trials = 2
+trials = 3
 
 returns = {"env": [], "lr": [], "pseudoreward": [], "episode": [], "return": []}
 episode_ixs = [i + 1 for i in range(episodes)]
@@ -46,7 +46,7 @@ if test_run == True:
     #df.lr = df.lr.astype('category')
     #df.lr.cat.remove_unused_categories() # make sure we don't plot extraneous learning rates
     # seaborn provides bootstrapped confidence intervals
-    seaborn.lineplot(x = "episode", y = "return", data = df, hue = "lr", style = 'use_entropy', legend = 'full')
+    seaborn.lineplot(x = "episode", y = "return", data = df, hue = "lr", style = 'pseudoreward', legend = 'full')
     plt.show()
 
 ### experiment

### 2019-03-19 writeup/paper.tex
fixed mountain car premature termination; added information content pseudoreward; general cleanup
added short corridor gridworld environment and updated paper

@@
 
-\title{CMPUT 609 Course Project}
+\title{Entropy and Information Content as Exploration Bonuses in REINFORCE}
 \author{Alan Chan}
@@
   \section{Introduction}
-  Policy gradient methods in Reinforcement Learning (RL) are a way of learning a policy in a Markov Decision Process (MDP) so as to maximize an objective function, usually the expected reward.
+  Policy gradient methods \citep[Chapter~13]{sutton2018reinforcement} in Reinforcement Learning (RL) are a way of learning a parametrized policy $\pi_\theta(a \mid s)$ in a Markov Decision Process (MDP) so as to maximize an objective function, usually the expected return $v_{\pi_\theta}(s_0)$ starting from a state $s_0$, where $v_{\pi_\theta}$ is the value function of $\pi_\theta$.
+In the following, we will assume that our state space $S$ and action space $A$ are both finite.
 
-  \section{Methods}
+  A key component of policy gradient methods, and RL techniques in general, is sufficient exploration of the state-action space to identify optimal action(s) for a given state.
+One possible technique to improve such exploration is the use of entropy or information content as a pseudoreward signal.
+More formally, let $R(s, a)$ be a random variable denoting the reward obtained from following action $a$ in state $s$.
+We define the following two quantities:
 
-  \section{Results & Discussion}
+  \begin{align*}
+    R_e(s, a) &:= R(s, a) - \sum_{a' \in A} \pi_\theta(a' \mid s) \ln\left(\pi_\theta(a' \mid s)\right),
+    R_i(s, a) &:= R(s, a) - \ln\left(\pi_\theta(a \mid s)\right),
+  \end{align*}
+
+  where $R_e(s, a)$ will be denoted the \textit{entropy pseudoreward}, and $R_i(s, a)$ will be denoted the \textit{information content pseudoreward}.
+Intuitively, for a given state-action pair $(s, a)$, $R_e(s, a)$ is maximized as a function of the policy $\pi_\theta$ when $\pi_\theta$ is a uniform distribution over all actions $a$.
+Thus, the reward is highest for a policy that uniformly selects actions at random, corresponding to a high degree of exploration.
+For a given state $s$, the maximum of $R_i(s, a)$ as a function of $a$ corresponds to a tradeoff between the maximum of $R(s, a)$ as a function of $a$ and $- \ln\left(\pi_\theta(a \mid s)\right)$ as a function of $a$.
+Intuitively, $- \ln\left(\pi_\theta(a \mid s)\right)$, the information content of taking action $a$, is at a maximum when $\pi_\theta(a \mid s)$ is at a minimum.
+In other words, the information content term in $R_i(s, a)$ encourages the agent to take actions that have low probability under the current policy, again corresponding to an increase in exploration.
+
+  In this report, we focus upon the use of the entropy and information content pseudorowards in the REINFORCE algorithm \citep{williams1992simple}, pseudocode of which is in \citet[p.~328]{sutton2018reinforcement}.
+We evaluate compare both of the pseudorewards with the vanilla reward $R(s, a)$ to determine if any of the pseudorewards facilitate better exploration of the state-action space and thus whether a higher return is achieved in less time.
+
+  \section{Experimental Details}
+  \subsection{Algorithms}
+  All code is available at \url{}.
+
+  REINFORCE is the main algorithm used for this comparison.
+We chose REINFORCE because of its relative simplicity in the class of policy gradient algorithms.
+We chose not to use a baseline, as is common in practice to reduce the variance of REINFORCE updates, so that we may isolate the effects of the different reward functions.
+
+  All methods were trained with SGD with step-sizes in the set $\{1 \cdot 10^{-4} \cdot 2^i : i \in \{0, 1, 2, 3, 4\}\}$.
+All algorithms were run for a set number of trials depending upon the environment, and the random seed was reset to the same value at the beginning of each set of trials for a given algorithm and environment to ensure fair comparison amongst all the algorithms.
+
+  We plot learning curves of all the methods for each of our environments.
+
+  \subsection{Environments}
+  We performed experiments on three environments. The first environment is the short corridor gridworld from \citet[p.~323]{sutton2018reinforcement}.
+
+  The second environment is cart pole,
+
+  The third environment is mountain car
+
+  \section{Results \& Discussion}
 
   \section{Conclusion}
+
+  \bibliography{bibliography}
+  \bibliographystyle{plain}
 \end{document}

### 2019-03-19 algs.py

@@ -16,8 +16,12 @@ class Policy(nn.Module):
 def train(env, lr, gamma, pseudoreward, episodes = 100, episode_length = None, render_env = False, print_return = False):
     # do REINFORCE because we only have one step-size
     # can do actor-critic later if there is time
-    action_dim = env.action_space.n
-    obs_dim = env.observation_space.shape[0]
+    try: # if we are working with gym
+        action_dim = env.action_space.n
+        obs_dim = env.observation_space.shape[0]
+    except: # if we are working with our own environments
+        action_dim = env.action_dim
+        obs_dim = env.obs_dim
     #print(action_dim, obs_dim)
     policy = Policy(obs_dim, action_dim).to(device)
     sgd = torch.optim.SGD(policy.parameters(), lr = lr)

### 2019-03-19 main.py

@@ -5,6 +5,7 @@ env1 = gym.make('CartPole-v0')
 env2 = gym.make('MountainCar-v0').env
 env3 = gym.make("Acrobot-v1").env
 env4 = gym.make("LunarLander-v2")
+env5 = SCGrid()
 
 # env1 and env4 have episodes that are short length so don't have to artificially terminate them
 
@@ -14,18 +15,18 @@ envs = [env1]
 pseudorewards = ["none", "entropy", "information_content"]
 
 run_experiment = False
-render_env = True
-print_return = True
+render_env = False
+print_return = False
 load_data = False
 test_run = not(run_experiment)
 save_data = run_experiment
-episodes = 3
+episodes = 10
 gamma = 0.99
-episode_length = 500
+episode_length = None
 
 # learning rates we will try
 lrs = [1e-4 * (2 ** i) for i in range(5)]
-trials = 3
+trials = 100
 
 returns = {"env": [], "lr": [], "pseudoreward": [], "episode": [], "return": []}
 episode_ixs = [i + 1 for i in range(episodes)]
@@ -37,7 +38,7 @@ if test_run == True:
             print("trials for pseudoreward = {}".format(pseudoreward))
             for trial in range(trials):
                 print("Starting trial {}".format(trial + 1))
-                data = train(env = env2, lr = lr, gamma = gamma, pseudoreward = pseudoreward, episodes = episodes, render_env = render_env, print_return = print_return, episode_length = episode_length)
+                data = train(env = env5, lr = lr, gamma = gamma, pseudoreward = pseudoreward, episodes = episodes, render_env = render_env, print_return = print_return, episode_length = episode_length)
                 df["pseudoreward"] += [pseudoreward] * episodes
                 df["return"] += data
                 df["episode"] += episode_ixs

### 2019-03-19 test_envs.py

@@ -1,1 +1,29 @@
 from includes import *
+
+class SCGrid():
+    def __init__(self):
+        self.reward = -1
+        self.actions = [-1, 1] # convenience used to index actions
+        self.action_dim = 2
+        self.obs_dim = 1
+        self.obs = [1] # all states look the same in this env, so only return one number
+        self.states = [0, 1, 2, 3]
+        self.curr_state = 0
+
+    def step(self, action):
+        # if in second state, reverse moves
+        if self.curr_state == 1:
+            action = 1 - action
+            self.curr_state += self.actions[action]
+        elif (self.curr_state + self.actions[action]) >= 0 and (self.curr_state + self.actions[action]) <= 3:
+            self.curr_state += self.actions[action]
+        if self.curr_state == 3:
+            done = True
+        else:
+            done = False
+        info = None
+        return self.obs, self.reward, done, info
+
+    def reset(self):
+        self.curr_state = 0
+        return self.obs

### 2019-03-30 main.py
experimental data and fixed some bugs in experiment

@@ -11,22 +11,22 @@ env5 = SCGrid()
 
 # consider artificially terminating episodes if they run too long?
 
-envs = [env1]
+envs = [env1, env5]
 pseudorewards = ["none", "entropy", "information_content"]
 
-run_experiment = False
+run_experiment = True
 render_env = False
 print_return = False
 load_data = False
-test_run = not(run_experiment)
+test_run = False
 save_data = run_experiment
-episodes = 10
+episodes = 50
 gamma = 0.99
 episode_length = None
 
 # learning rates we will try
-lrs = [1e-4 * (2 ** i) for i in range(5)]
-trials = 100
+lrs = [1e-4 * (2 ** i) for i in range(4)]
+trials = 200
 
 returns = {"env": [], "lr": [], "pseudoreward": [], "episode": [], "return": []}
 episode_ixs = [i + 1 for i in range(episodes)]
@@ -57,8 +57,11 @@ if run_experiment == True:
             for pseudoreward in pseudorewards:
                 np.random.seed(609) # the random seed should be reset for every type of experiment since we want fair comparisons between algs
                 t_start = time.time()
-                env_name = env.unwrapped.spec.id
-                print("performing trials for env = {}, lr = {}, use_entropy = {}".format(env_name, lr, use_entropy))
+                try:
+                    env_name = env.unwrapped.spec.id
+                except:
+                    env_name = env.name
+                print("performing trials for env = {}, lr = {}, pseudoreward = {}".format(env_name, lr, pseudoreward))
                 for trial in range(trials):
                     returns["env"] += [env_name] * episodes
                     returns["lr"] += [lr] * episodes
@@ -100,10 +103,13 @@ if save_data is True:
 if load_data is True or run_experiment is True:
     for env in envs:
         for pseudoreward in pseudorewards:
-            env_name = env.unwrapped.spec.id
+            try:
+                env_name = env.unwrapped.spec.id
+            except:
+                env_name = env.name
             curr_env = df.env == env_name
-            curr_entropy = df.use_entropy == use_entropy
-            ax = seaborn.lineplot(x = "episode", y = "return", hue = "lr", legend = "full", data = df.loc[curr_env & curr_entropy, :])
+            curr_pseudoreward = df.pseudoreward == pseudoreward
+            ax = seaborn.lineplot(x = "episode", y = "return", hue = "lr", legend = "full", data = df.loc[curr_env & curr_pseudoreward, :])
[…]

### 2019-03-30 writeup/paper.tex
updated paper

--- 2019-03-06 writeup/paper.tex
@@
 
-\usepackage[autostyle, english = american]{csquotes}
-\MakeOuterQuote{"}
+%\usepackage[autostyle, english = american]{csquotes}
+%\MakeOuterQuote{"}
 \usepackage[utf8]{inputenc}
@@
   \section{Introduction}
-  Policy gradient methods in Reinforcement Learning (RL) are a way of learning a policy in a Markov Decision Process (MDP) so as to maximize an objective function, usually the expected reward.
+  Policy gradient methods \citep[Chapter~13]{sutton2018reinforcement} in Reinforcement Learning (RL) are a way of learning a parametrized policy $\pi_\theta(a \mid s)$ in a Markov Decision Process (MDP) \citep[Chapter~3]{sutton2018reinforcement} so as to maximize an objective function, usually the expected return $v_{\pi_\theta}(s_0)$ starting from a state $s_0$, where $v_{\pi_\theta}$ is the value function of $\pi_\theta$.
+More concretely,
 
-  \section{Methods}
+  \begin{equation*}
+    v_{\pi_\theta}(s_0) := \Ex\left[G_t \mid S_t = s_0\right] = \Ex\left[\sum_{t' = t}^T \gamma^{t'- t} R(S_t, A_t) \right],
+  \end{equation*}
 
-  \section{Results & Discussion}
+  where $\gamma$ is the discount factor and $T$ is our termination time. $T$ itself is a random variable, and may be $\infty$ in the case of a continuing MDP. $G_t$ is called the return at time $t$.
+
+  In the following, we will assume that our state space $\mathcal{S}$ and action space $\mathcal{A}$ are both finite.
+
+  A key component of policy gradient methods, and RL techniques in general, is sufficient exploration of the state-action space to identify actions for a given state that will lead to high return.
+One possible technique to improve such exploration is the use of entropy or information content as a reward in addition to the reward $R(s, a)$ from the external environment.
+More formally, let $R(a, a)$ be a random variable denoting the reward obtained from following action $a$ in state $s$.
+We define the following two quantities:
+
+  \begin{align*}
+    R_e(s, a) &:= R(s, a) - \sum_{a' \in A} \pi_\theta(a' \mid s) \ln\left(\pi_\theta(a' \mid s)\right),\\
+    R_i(s, a) &:= R(s, a) - \ln\left(\pi_\theta(a \mid s)\right),
+  \end{align*}
+
+  where $R_e(s, a)$ will be denoted the \textit{entropy pseudoreward}, and $R_i(s, a)$ will be denoted the \textit{information content pseudoreward}.
+We will use the term pseudoreward to refer to reward signals that are not solely from the environment.
+The reward $R(s, a)$ that comes from the environment will just be referred to as the reward.
+Intuitively, for a given state-action pair $(s, a)$, $R_e(s, a)$ is maximized as a function of the policy $\pi_\theta$ when $\pi_\theta$ is a uniform distribution over all actions $a$.
+Thus, the reward is highest for a policy that uniformly selects actions at random, corresponding to a high degree of exploration.
+For a given state $s$, the maximum of $R_i(s, a)$ as a function of $a$ corresponds to a tradeoff between the maximum of $R(s, a)$ as a function of $a$ and $- \ln\left(\pi_\theta(a \mid s)\right)$ as a function of $a$.
+Intuitively, $- \ln\left(\pi_\theta(a \mid s)\right)$, the information content of taking action $a$, is at a maximum when $\pi_\theta(a \mid s)$ is at a minimum.
+In other words, the information content term in $R_i(s, a)$ encourages the agent to take actions that have low probability under the current policy, again corresponding to an increase in exploration.
+
+  In this report, we focus upon the use of the entropy and information content pseudorowards in the REINFORCE algorithm \citep{williams1992simple}.
+We evaluate compare both of the pseudorewards with the reward $R(s, a)$ to determine if any of the pseudorewards facilitate better exploration of the state-action space and thus whether a higher return is achieved in less time.
+
+  \section{Experimental Details}
+  \subsection{Algorithms}
+  All code is available at \url{https://github.com/SwordShieldMouse/cmput-609-project}.
+
+  REINFORCE is the main, overarching algorithm used for the comparison of pseudoreward functions.
+We chose REINFORCE because of its relative simplicity in the class of policy gradient algorithms.
+We chose not to use a baseline, as is common in practice to reduce the variance of REINFORCE updates, so that we may isolate the effects of the different reward functions.
+
+  We use the same functional form for the policy $\pi(\cdot \mid \cdot, \bm{\theta})$ for all of our experiments, where $\bm{\theta}$ is the parameter vector.
+We let $\bm{\theta} \in \mathbb{R}^{|\mathcal{S}| \times |\mathcal{A}|}$, so we have
+
+  \begin{equation*}
+    \pi(\cdot \mid S, \bm{\theta}) :=  Softmax(S^T \bm{\theta}),
+  \end{equation*}
+
+  where we slightly abuse notation and write $S$ for the features of state $S$.
+
+  As noted above, we will focus our comparison of the reward $R(s, a)$ with two types of pseudorewards: the entropy pseudoreward $R_e(s, a)$ and the information content pseudoreward $R_i(s, a)$.
+There are thus three alternatives we will test.
+1: REINFORCE using $R(s, a)$ without any additional pseudoreward signal \ref{alg:vanilla}; 2: REINFORCE with $R_i(s, a)$ replacing the reward \ref{alg:information_content}; 3: REINFORCE with $R_e(s, a)$ replacing the reward \ref{alg:entropy}.
+Of course, during evaluation of the algorithms, we will use $R(s, a)$ in calculating the return for each algorithm.
+
+  \begin{algorithm}
+    \caption{REINFORCE with $R(s, a)$. Base pseudocode from \citet[p.328]{sutton2018reinforcement}.}
+    \label{alg:vanilla}
+    \begin{algorithmic}
+      \State Input: policy $\pi(a \mid s, \bm{\theta})$
+      \State Parameter: step-size $\alpha$
+      \For{each episode}
+        \State Generate an episode $\{(S_i, A_i, R_{i + 1})\}_{i = 0}^{T - 1}$ following $\pi(\cdot \mid \cdot, \bm{\theta})$
+        \For{$t = 0, \cdots, T - 1$}
+          \State $G := \sum_{k = t + 1}^T \gamma^{k - t - 1} R_k$
+          \State $\bm{\theta} := \alpha \gamma^t G \nabla \ln \pi(A_t \mid S_t, \bm{\theta})$.
+        \EndFor
+      \EndFor
+    \end{algorithmic}
+  \end{algorithm}
+
+  \begin{algorithm}
+    \caption{REINFORCE with $R_e(s, a)$. Base pseudocode from \citet[p.328]{sutton2018reinforcement}.}
+    \label{alg:entropy}
+    \begin{algorithmic}
+      \State Input: policy $\pi(a \mid s, \bm{\theta})$
+      \State Parameter: step-size $\alpha$
+      \For{each episode}
+        \State Generate an episode $\{(S_i, A_i, R_{i + 1})\}_{i = 0}^{T - 1}$ following $\pi(\cdot \mid \cdot, \bm{\theta})$
+        \For{$t = 0, \cdots, T - 1$}
+          \State $G := \sum_{k = t + 1}^T \gamma^{k - t - 1} (R_k - \sum_{a \in \mathcal{A}} \pi(a \mid S_{k - 1}, \bm{\theta}) \ln \pi(a \mid S_{k - 1}, \bm{\theta}))$
+          \State $\bm{\theta} := \alpha \gamma^t G \nabla \ln \pi(A_t \mid S_t, \bm{\theta})$.
+        \EndFor
+      \EndFor
+    \end{algorithmic}
+  \end{algorithm}
+
+  \begin{algorithm}
+    \caption{REINFORCE with $R_i(s, a)$. Base pseudocode from \citet[p.328]{sutton2018reinforcement}.}
+    \label{alg:information_content}
+    \begin{algorithmic}
+      \State Input: policy $\pi(a \mid s, \bm{\theta})$
+      \State Parameter: step-size $\alpha$
+      \For{each episode}
+        \State Generate an episode $\{(S_i, A_i, R_{i + 1})\}_{i = 0}^{T - 1}$ following $\pi(\cdot \mid \cdot, \bm{\theta})$
+        \For{$t = 0, \cdots, T - 1$}
+          \State $G := \sum_{k = t + 1}^T \gamma^{k - t - 1} (R_k - \ln \pi(A_{k - 1} \mid S_{k - 1}, \bm{\theta}))$
+          \State $\bm{\theta} := \alpha \gamma^t G \nabla \ln \pi(A_t \mid S_t, \bm{\theta})$.
+        \EndFor
+      \EndFor
+    \end{algorithmic}
+  \end{algorithm}
+
+  All methods were trained with SGD with step-sizes in the set $\{1 \cdot 10^{-4} \cdot 2^i : i \in \{0, 1, 2, 3, 4, 5\}\}$.
+All algorithms were run for a set number of trials depending upon the environment, and the random seed was reset to the same value (609) at the beginning of each set of trials for a given algorithm and environment to ensure fair comparison amongst all the algorithms.
+
+  We plot learning curves of all the methods for each of our environments.
+
+  All experiments were run in Python 3.7 on an early 2015 MacBook Pro running macOS Mojave 10.14. PyTorch was used for implementing the policy architecture and calculating gradients.
+
+  \subsection{Environments}
+  For all environments, the discount factor was set to $\gamma := 0.99$.
+
+  We performed experiments on three environments.
+The first environment is the short corridor gridworld from \citet[p.~323]{sutton2018reinforcement}.
+There are four states, one of which is terminal (the rightmost state).
+All the states are represented by the same number.
+In this case, we chose 1.
+At each of the three non-terminal states, there are two actions: right and left.
+In the leftmost state and the third state from the left, the actions have the usual consequences of moving the agent right or left if possible.
+In the second state from the left, the results of the actions are switched, so that right moves left and left moves right.
+Here, we set the number of episodes to be $100$ and the number of trials to be $500$.
+We chose the short corridor grid world to compare our results to those in the book.
+
+  The second environment is cart pole, where the number of episodes was $500$ and the number of trials was $500$.
+The goal of the agent is to balance a pole situated on a cart by applying force to move the cart.
+
+  The third environment is mountain car, where we ran experiments for $10$ episodes and $100$ trials.
+The comparatively low number of episodes and trials is due to the difficulty of the mountain car environment.
+
+  \section{Results \& Discussion}
 
   \section{Conclusion}
+
+  \bibliography{bibliography}
+  \bibliographystyle{plain}
 \end{document}

### 2019-03-30 main.py
added more graphing code

@@ -14,10 +14,10 @@ env5 = SCGrid()
 envs = [env1, env5]
 pseudorewards = ["none", "entropy", "information_content"]
 
-run_experiment = True
+run_experiment = False
 render_env = False
 print_return = False
-load_data = False
+load_data = True
 test_run = False
 save_data = run_experiment
 episodes = 50
@@ -109,11 +109,30 @@ if load_data is True or run_experiment is True:
                 env_name = env.name
             curr_env = df.env == env_name
             curr_pseudoreward = df.pseudoreward == pseudoreward
+            plt.figure(figsize = (15, 10))
             ax = seaborn.lineplot(x = "episode", y = "return", hue = "lr", legend = "full", data = df.loc[curr_env & curr_pseudoreward, :])
             ax.set_title("Learning curves for pseudoreward = {} on {}".format(pseudoreward, env_name))
-            plt.savefig("figs\\{}-{}-experiments.png".format(env_name, pseudoreward))
-            plt.show()
+            plt.savefig("figs\\{}-{}.png".format(env_name, pseudoreward))
+            #plt.show()
 
 
 ## do statistical tests?
-# other ways of evaluating: episodes required to get to a certain reward
+# plot best learning curves from each method on one plot
+for env in envs:
+    try:
+        env_name = env.unwrapped.spec.id
+    except:
+        env_name = env.name
+    curr_env = df.env == env_name
+    best_none = (df.pseudoreward == "none") & (df.lr == 8e-4)
+    if env_name == "ShortCorridorGridworld":
+        best_entropy = (df.pseudoreward == "entropy") & (df.lr == 1e-4)
+    else:
+        best_entropy = (df.pseudoreward == "entropy") & (df.lr == 8e-4)
+    best_info = (df.pseudoreward == "information_content") & (df.lr == 8e-4)
+    rows = curr_env & (best_none | best_entropy | best_info)
+    plt.figure(figsize = (15, 10))
+    ax = seaborn.lineplot(x = "episode", y = "return", style = "lr", hue = "pseudoreward", legend = "full", data = df.loc[rows, :])
+    ax.set_title("Comparison of algorithms on env = {}".format(env_name))
+    plt.savefig("figs\\{}-compare.png".format(env_name))
+    #plt.show()

### 2019-04-03 writeup/paper.tex
fixed random seed issue and taking derivative of pseudoreward

@@
 \usepackage{caption}
+\usepackage{subcaption}
 
@@
 
+\captionsetup[subfigure]{subrefformat=simple,labelformat=simple}
+  \renewcommand\thesubfigure{(\alph{subfigure})}
+
+  \renewcommand{\thesubfigure}{(\alph{subfigure})}
+\captionsetup[sub]{labelformat=simple}
+
 \title{Entropy and Information Content as Exploration Bonuses in REINFORCE}
@@
 
+  \begin{abstract}
+
+  \end{abstract}
+
   \section{Introduction}
-  Policy gradient methods \citep[Chapter~13]{sutton2018reinforcement} in Reinforcement Learning (RL) are a way of learning a parametrized policy $\pi_\theta(a \mid s)$ in a Markov Decision Process (MDP) \citep[Chapter~3]{sutton2018reinforcement} so as to maximize an objective function, usually the expected return $v_{\pi_\theta}(s_0)$ starting from a state $s_0$, where $v_{\pi_\theta}$ is the value function of $\pi_\theta$.
+  Policy gradient methods \citep[Chapter~13]{sutton2018reinforcement} in Reinforcement Learning (RL) are a way of learning a parametrized policy $\pi_\theta(a \mid s)$ in a Markov Decision Process (MDP) \citep[Chapter~3]{sutton2018reinforcement} so as to maximize an objective function.
+The objective function in a policy gradient setting is usually the expected return $v_{\pi_\theta}(s_0)$ starting from a state $s_0$, where $v_{\pi_\theta}$ is the value function of $\pi_\theta$.
 More concretely,
@@
   A key component of policy gradient methods, and RL techniques in general, is sufficient exploration of the state-action space to identify actions for a given state that will lead to high return.
-One possible technique to improve such exploration is the use of entropy or information content as a reward in addition to the reward $R(s, a)$ from the external environment.
+One possible technique to improve such exploration is the use of entropy or information content as a reward bonus in addition to the reward $R(s, a)$ from the external environment.
 More formally, let $R(a, a)$ be a random variable denoting the reward obtained from following action $a$ in state $s$.
@@
 We evaluate compare both of the pseudorewards with the reward $R(s, a)$ to determine if any of the pseudorewards facilitate better exploration of the state-action space and thus whether a higher return is achieved in less time.
+We hypothesize that both of the reward bonuses will substantially improve exploration and enable agents to accumulate more return sooner than without the bonuses.
 
@@
 
-  where we slightly abuse notation and write $S$ for the features of state $S$.
+  where we slightly abuse notation and write $S$ for the features of state $S$. For every single trial, the initial weight vector was set to the same random value.
 
@@
 There are thus three alternatives we will test.
-1: REINFORCE using $R(s, a)$ without any additional pseudoreward signal \ref{alg:vanilla}; 2: REINFORCE with $R_i(s, a)$ replacing the reward \ref{alg:information_content}; 3: REINFORCE with $R_e(s, a)$ replacing the reward \ref{alg:entropy}.
+1: REINFORCE using $R(s, a)$ without any additional pseudoreward signal \cref{alg:vanilla} (we will sometimes refer to this as the "vanilla" reward); 2: REINFORCE with $R_i(s, a)$ replacing the reward \cref{alg:information_content}; 3: REINFORCE with $R_e(s, a)$ replacing the reward \cref{alg:entropy}.
 Of course, during evaluation of the algorithms, we will use $R(s, a)$ in calculating the return for each algorithm.
@@
 
-  All methods were trained with SGD with step-sizes in the set $\{1 \cdot 10^{-4} \cdot 2^i : i \in \{0, 1, 2, 3, 4, 5\}\}$.
+  All methods were trained with SGD with step-sizes in the set $\{1 \cdot 10^{-4} \cdot 2^i : i \in \{0, 1, 2, 3, 4\}\}$.
 All algorithms were run for a set number of trials depending upon the environment, and the random seed was reset to the same value (609) at the beginning of each set of trials for a given algorithm and environment to ensure fair comparison amongst all the algorithms.
@@
 
-  We performed experiments on three environments.
+  We performed experiments on two environments.
 The first environment is the short corridor gridworld from \citet[p.~323]{sutton2018reinforcement}.
@@
 In the second state from the left, the results of the actions are switched, so that right moves left and left moves right.
-Here, we set the number of episodes to be $100$ and the number of trials to be $500$.
-We chose the short corridor grid world to compare our results to those in the book.
 
-  The second environment is cart pole, where the number of episodes was $500$ and the number of trials was $500$.
+  The second environment is cart pole.
 The goal of the agent is to balance a pole situated on a cart by applying force to move the cart.
+Cart pole was chosen also for its relative simplicity, but also for the fact that it represents a control scenario.
 
-  The third environment is mountain car, where we ran experiments for $10$ episodes and $100$ trials.
-The comparatively low number of episodes and trials is due to the difficulty of the mountain car environment.
+  For all environments, the number of episodes was set to $100$ and the number of trials to be $300$.
+Given limited computational resources, for all episodes the max length of the episode was set to be 500.
 
-  \section{Results \& Discussion}
+  \section{Results}
+  In the plots, the solid line is the average of all of the returns across all trials for a given episode, while the shaded region gives the $95\%$ confidence intervals from the standard error.
+
+  Each plot in \cref{fig:cartpole,fig:shortcorridorgridworld} is for a given algorithm and environment across the range of step-sizes.
+In general, except for the entropy pseudoreward algorithm on the short corridor gridworld, the claim that a learning rate of $8 \cdot 10^{-4}$ yielded the best performance out of the all the learning rates is most consistent with the data.
+
+  \begin{figure}
+    \begin{subfigure}{\textwidth}
+      \centering
+      \includegraphics[width=\linewidth]{../figs/CartPole-v0-none}
+      \label{fig:cartpole-none}
+    \end{subfigure}
+    \begin{subfigure}{\textwidth}
+      \centering
+      \includegraphics[width=\linewidth]{../figs/CartPole-v0-entropy}
+      \label{fig:cartpole-entropy}
+    \end{subfigure}
+    \begin{subfigure}{\textwidth}
+      \centering
+      \includegraphics[width=\linewidth]{../figs/CartPole-v0-information_content}
+      \label{fig:cartpole-information-content}
+    \end{subfigure}
+    \caption{Cart pole experiments.}
+    \label{fig:cartpole}
+  \end{figure}
+
+  \begin{figure}
+    \begin{subfigure}{\textwidth}
+      \centering
+      \includegraphics[width=\linewidth]{../figs/ShortCorridorGridworld-none}
+      \label{fig:shortcorridorgridworld-none}
+    \end{subfigure}
+    \begin{subfigure}{\textwidth}
+      \centering
+      \includegraphics[width=\linewidth]{../figs/ShortCorridorGridworld-entropy}
+      \label{fig:shortcorridorgridworld-entropy}
+    \end{subfigure}
+    \begin{subfigure}{\textwidth}
+      \centering
+      \includegraphics[width=\linewidth]{../figs/ShortCorridorGridworld-information_content}
+      \label{fig:shortcorridorgridworld-information-content}
+    \end{subfigure}
+    \caption{Short corridor gridworld experiments.}
+    \label{fig:shortcorridorgridworld}
+  \end{figure}
+
+  In \cref{fig:compare}, we compare the learning curves of the algorithms, selecting a learning rate for each, for a given environment.
+The learning rate selected corresponded to the curve that attained the highest return most of the time in \cref{fig:cartpole,fig:shortcorridorgridworld}, which was $8 \cdot 10^{-4}$ for most algorithm/environment/learning rate combinations, except for the entropy pseudoreward algorithm on the short corridor gridworld, as noted above.
+This selection was done visually.
+
+  \begin{figure}
+    \centering
+    \begin{subfigure}{\textwidth}
+      \centering
+      \includegraphics[width = \linewidth]{../figs/CartPole-v0-compare}
+      \label{fig:compare-cartpole}
+    \end{subfigure}
+    \begin{subfigure}{\textwidth}
+      \centering
+      \includegraphics[width = \linewidth]{../figs/ShortCorridorGridworld-compare}
+      \label{fig:compare-shortcorridorgridworld}
+    \end{subfigure}
+    \caption{Comparison of learning curves across different algorithms. }
+    \label{fig:compare}
+  \end{figure}
+
+  \section{Discussion}
+  It is interesting to note that the entropy pseudoreward algorithm performed the worst on the short corridor gridworld.
+In fact, performance got worse as time passed, and the average return drastically decreased with higher step-sizes, in contrast to all of the other algorithm and environment combinations.
+On the other hand, the information content pseudoreward performed similarly to the "vanilla" reward algorithm.
+In fact, both vanilla reward and information content pseudoreward algorithms approached the optimal return of -11.6 \citep[p.~323]{sutton2018reinforcement} with increasing episodes.
+
+  A low return on the short corridor gridworld means that a lot of time was spent in the non-terminal states.
+This is likely to happen if a agent selects one action with probability much higher than selecting the other probability so that the agent oscillates between the state where the actions are switched and another non-terminal state.
+
+  In \ref{fig:compare-cartpole}, from episodes 0 to around 40, the pseudoreward algorithms seem to begin to outperform the vanilla reward algorithm.
+However, from episode 40 onwards, the slope of the learning curves for the entropy and information content pseudorewards seems to decrease, while the slope for the learning curve of the vanilla reward continues to increase.
+This result is consistent with the idea that the entropy and information content pseudorewards incentivize exploration of the state-action space by encouraging, respectively, high entropy policies and taking actions with low probability.
+However, as time progresses and as more of the state-action space is explored, such exploration might be detrimental if an optimal policy is already found.
+To that extent, entropy and information content pseudorewards act similarly to $\epsilon$-greedy methods in leraning optimal value functions.
+This result suggests an annealing procedure similar to that used for $\epsilon$-greedy methods, where we gradually reduce the additional entropy or information content term to zero more episodes are experienced.
+
+  Additionally, the entropy pseudoreward seems to suffer more from this detrimental exploration effect than the information content pseudoreward.
+The slope of the information content pseudoreward learning curve does not decrease as steeply as the slope for the entropy pseudoreward learning curve.
+It is possible that the entropy pseudorewards encourages exploration to a greater extent than the information content pseudoreward.
+In the first case, policies with higher entropy are encouraged.
+In the second case, the agent is encouraged to select actions with low action probability.
+While these two situations are related, a higher entropy policy may require probability density to be evenly distributed across all outcomes.
+If the optimal policy has low entropy, that policy may not be the optimal policy under the entropy pseudoreward setting.
+In contrast, the information content pseudoreward is a less stringent, more local requirement.
+If a low probability action is taken, the reward gained is relatively high.
+The reward bonus depends upon the action actually taken.
+If the optimal policy has a low probability for a given action, the probability of taking that action may be increased rarely in the information content setting since the probability of taking that action in the first place to receive the information content bonus is low.
+Hence, even if the optimal policy has actions with low probability, the optimal policy is not itself impossible to reach.
+The learned policy might oscillate in a neighbourhood of the optimal policy, but the situation is better than with an entropy pseudoreward.
+
+  In \ref{fig:compare-shortcorridorgridworld}, there is essentially little difference in the learning curves for vanilla reward and information content.
+Hence, assuming that the agent has the optimal policy The difference between both entropy pseudoreward and both information content pseudoreward and vanilla reward is significant, as already noted above.
 
   \section{Conclusion}
+  In the cart pole environment, using entropy and information content as reward bonuses seems to improve the returns accumulated in early episodes.
+However, this advantage is soon lost as exploration prevents convergence to the optimal policy as the environment is sufficiently explored.
+
+  In the short corridor gridworld,
+
+  Our results suggest that the benefits of using entropy and information content as exploration bonuses depend upon the environment.
+While there may be a benefit to using these bonuses for episodes near the start, these bonuses may hinder convergence to an optimal policy.
+However, such characteristics may be beneficial in non-stationary environments and should be explored in future experiments.
 

### 2019-04-03 algs.py

@@ -25,7 +25,9 @@ def train(env, lr, gamma, pseudoreward, episodes = 100, episode_length = None, r
     #print(action_dim, obs_dim)
     policy = Policy(obs_dim, action_dim).to(device)
     sgd = torch.optim.SGD(policy.parameters(), lr = lr)
-
+    #false_rewards = {"episode": [], "false_reward": []}
+    #policy_probs = {"timestep": [], "prob": []}
+    #plt.figure(figsize = (15, 10))
     returns = []
     for episode in range(episodes):
         obs = env.reset()
@@ -34,12 +36,14 @@ def train(env, lr, gamma, pseudoreward, episodes = 100, episode_length = None, r
         # make sure we use the actual rewards to evaluate how we do
         actual_rewards = []
 
+        #false_rewards = []
+
         # stores the rewards to which we might add entropy
         rewards = []
 
         states = [torch.Tensor(obs).to(device)]
         actions = []
-        t = 0 # to index episode time
+        t = 0 # to index episode timestep
         while done is not True:
             if render_env is True:
                 env.render()
@@ -48,6 +52,13 @@ def train(env, lr, gamma, pseudoreward, episodes = 100, episode_length = None, r
             m = Categorical(logits = log_probs)
             #print("time = {}".format(t), log_probs, obs)
             action = m.sample()
+
+            # record the action probs
+            """policy_probs["timestep"].append(t)
+            if action == 0: # records prob of going left
+                policy_probs["prob"].append(np.exp(m.log_prob(action).detach().item()))
+            else:
+                policy_probs["prob"].append(1 - np.exp(m.log_prob(action).detach().item()))"""
 
             obs, reward, done, info = env.step(action.item())
 
@@ -59,9 +70,16 @@ def train(env, lr, gamma, pseudoreward, episodes = 100, episode_length = None, r
 
             # adding pseudorewards possibly to help with exploration
             if pseudoreward == "entropy":
-                reward += m.entropy()
+                reward += m.entropy().detach() # detach so that we don't take the derivative of this
+                #print(m.entropy())
+                #false_rewards["episode"].append(episode)
+                #false_rewards["false_reward"].append(reward.detach().item())
+                #false_rewards.append(reward.detach().item())
             elif pseudoreward == "information_content":
-                reward -= m.log_prob(action)
+                reward -= m.log_prob(action).detach() # detach so that we don't take the derivative of this
[…]

### 2019-04-03 main.py

@@ -11,22 +11,22 @@ env5 = SCGrid()
 
 # consider artificially terminating episodes if they run too long?
 
-envs = [env1, env5]
-pseudorewards = ["none", "entropy", "information_content"]
+envs = [env5, env1]
+pseudorewards = ["entropy", "information_content", "none"]
 
-run_experiment = False
+run_experiment = True
 render_env = False
 print_return = False
-load_data = True
+load_data = False
 test_run = False
 save_data = run_experiment
-episodes = 50
+episodes = 100
 gamma = 0.99
-episode_length = None
+episode_length = 200
 
 # learning rates we will try
 lrs = [1e-4 * (2 ** i) for i in range(4)]
-trials = 200
+trials = 300
 
 returns = {"env": [], "lr": [], "pseudoreward": [], "episode": [], "return": []}
 episode_ixs = [i + 1 for i in range(episodes)]
@@ -56,6 +56,7 @@ if run_experiment == True:
         for lr in lrs:
             for pseudoreward in pseudorewards:
                 np.random.seed(609) # the random seed should be reset for every type of experiment since we want fair comparisons between algs
+                torch.manual_seed(609)
                 t_start = time.time()
                 try:
                     env_name = env.unwrapped.spec.id
@@ -69,12 +70,9 @@ if run_experiment == True:
 
                     trial_t_start = time.time()
 
-                    if env == env2 or env == env3:
-                        print("will limit episode length to {} time steps".format(episode_length))
-                        # if we are doing either of these environments, limit the episode length b/c they take a v long time to run
-                        data = train(env = env, lr = lr, gamma = gamma, pseudoreward = pseudoreward, episodes = episodes, episode_length = episode_length, render_env = render_env, print_return = print_return)
-                    else:
-                        data = train(env = env, lr = lr, gamma = gamma, pseudoreward = pseudoreward, episodes = episodes, render_env = render_env, print_return = print_return)
+                    #print("will limit episode length to {} time steps".format(episode_length))
+                    # if we are doing either of these environments, limit the episode length b/c they take a v long time to run
+                    data = train(env = env, lr = lr, gamma = gamma, pseudoreward = pseudoreward, episodes = episodes, episode_length = episode_length, render_env = render_env, print_return = print_return)
 
[…]
